FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-18 paper.tex
sec 3
delete GPU figure because we have no implementation for the arrow

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{microtype}

@@

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}
\newcommand{\GPUFPX}{GPU-FPX}
\newcommand{\FloatTracker}{FloatTracker}
\newcommand{\FT}{\FloatTracker}
\newcommand{\Fp}{Floating-point} % hyphen or no?
\newcommand{\fp}{floating-point} % hyphen or no?
\newcommand{\CSTG}{stack graph}
\newcommand{\CPP}{\code{C++}}
\newcommand{\Dendro}{\textsc{Dendro}}
\newcommand{\urlaccess}[2]{\url{#1}}
\newcommand{\Nan}{\code{NaN}}
\newcommand{\NaN}{\Nan}
\newcommand{\Inf}{\code{Inf}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zerocode}[1]{\zerowidth{\code{#1}}}

@@

\begin{figure}[t]\centering
  \includegraphics[trim=0 0 0 60,clip,width=0.95\columnwidth]{fig/real_vs_fp.png}
  \caption{Floats are spread across the real number line.}
  \label{f:real-vs-fp}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/genpropkill-outline.png}
  \caption{Gen, Prop, Kill: Lifetime of a \NaN{} value}
  \label{f:gpk}
\end{figure}

Unhandled exceptional values have a lifetime: they are born, or \emph{generated}, by some operation; they \emph{propagate} through other operations; and they either appear in the program output, go out of scope, or get \emph{killed} by a numeric operation.
\Cref{f:gpk} summarizes this \emph{gen}-\emph{prop}-\emph{kill} process.
The gens and props are straightforward; see above for examples~(\cref{s:exnvalue}).
The kills often arise from numeric comparisons (\code{<}, \code{=}, etc.),
but exponents (\code{1{\string^}NaN}) and over-eager matrix
optimizations~\cite{ddghlllprr-correctness-2022} can kill exceptions as well.

@@

\FlowFPX{} is a toolkit for tracking down \fp{} exceptions.
The two primary tools we cover in this paper are \FT{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.

@@

\FT{} is implemented as a Julia library, and is available for download on JuliaHub.\footnote{\url{https://juliahub.com/ui/Packages/FloatTracker/dBXig/1.0.0}}

@@

\FT{}'s first purpose is to monitor \fp{} operations for exceptional value lifetime events.
\FT{} interposes on arithmetic operations and standard library calls that involve a \fp{} value and logs a stack trace to that point whenever an exception is generated, propagated, or killed.
\FT{} can also log the arguments to the suspect call, so you can get \emph{normal, concrete} values to help discover what is going wrong.
A programmer can then use these stack traces to pinpoint where an exceptional value like \NaN{} came into being or where a guard against a \NaN{} kill is needed.

All the logs that \FT{} captures are straight from Julia's \texttt{stacktrace()} function.
\FT{} writes these logs to three files: one file for each kind of event in a \fp{} exception's lifetime.
This way, if a user of \FT{} is attempting to determine where a \NaN{} or an \Inf{} came from, they can look in the file for all the gen events.
Likewise, if the user wants to find where \NaN{}s get killed, then they can look in the kills file.

@@

The second role of \FT{} is to fuzz code from the inside out.
While can fuzz a program by giving it random inputs, this might not help you find \emph{all} the places where an exceptional value could crop up during a calculation.
Demmel et. al suggest fuzzing by injecting exceptional values at intermediate calculations;~\cite{ddghlllprr-correctness-2022} this is the approach that we enable with \FT{}.

\FT{}'s exception tracking infrastructure offers us a simple way to inject exceptional values wherever we want in our code: we are already interposing on \fp{} operations to monitor for exceptional lifetime values, so we can inject a \NaN{} at any of these points.

Since the injection happens at random, \FT{} also provides a way to save \NaN{} injections so that injections that exercise bugs can be replayed to examine the bug repeatedly and to verify that hardening efforts have indeed eliminated the vulnerability.

We have a global configuration system that allows us to turn on fuzzing and configure where, when, and how many \NaN{}s to inject, as well as whether or not to record or replay a recording:

\begin{itemize}
\item \texttt{replay::String} Name of a file to read an injection recording from. If set, this option makes all others irrelevant.
\item \texttt{record::String} Name of a file to write injection events to for later replay.
\item \texttt{active::Bool} Inject only if this is true.
\item \texttt{n\_inject::Int64} Inject at most this many \NaN{}s.
\item \texttt{odds::Int64} Inject if \texttt{rand(odds) == 1}.
\item \texttt{functions::Array} Inject if we are within the dynamic extent of any of these functions.
\item \texttt{libraries::Array} Inject if the top-most stack frame\footnote{After stripping off the stack frames that come from \FT{} itself.} comes from any of these libraries.
\end{itemize}

Then, each time we intercept a \fp{} operation, we check to see if we should inject a \NaN{}. First we check if we're replaying a recording: if we are, just do as the recording says. Otherwise, we check that the injector is \texttt{active}, we have \NaN{}s remaining to inject (\texttt{n\_inject}), and we roll an \texttt{odds}-sided die to see if we should inject. Finally, check the current stack trace against the lists of functions and libraries to inject in. If all looks good, we inject.

@@

While other tools like profilers might accomplish the tracking that \FT{} does by writing extensions to the language's compiler or runtime, \FT{} works entirely in user land.
\FT{} provides custom types---\texttt{TrackedFloat16}, \texttt{TrackedFloat32}, and \texttt{TrackedFloat64}---that wrap their corresponding \texttt{Float16}, \texttt{Float32}, and \texttt{Float64} counterparts.
\FT{} also defines methods for every function in the \texttt{Base} module.
This lets \FT{} intercept all \fp{} operations involving \texttt{TrackedFloat} types.

@@

This new method does a few things:

\begin{enumerate}
\item First it checks if it should inject a \NaN{} here.
\item If no \NaN{} is to be injected, it unwraps the \texttt{TrackedFloat}s to get to the primitive \texttt{Float} value, and computes the result normally.
\item It examines the arguments as well as the return value, and then uses \cref{fig:lifetime-class} to determine the type of event and logs it if it is of interest.
\item It wraps the result in a new \texttt{TrackedFloat} value.
\end{enumerate}

\begin{figure}[h]
  \centering
  \begin{tabular}{ccc}
    Arguments & Result & Event Type \\
    \hline
    No \NaN{} & \NaN{} & Gen \\
    \NaN{} & \NaN{} & Prop \\
    \NaN{} & No \NaN{} & Kill \\
  \end{tabular}
  \label{fig:lifetime-class}
  \caption{Exceptional value event classification rules}
\end{figure}

The same classification rules apply to \Inf{} and would apply to any other value of interest.

These new methods had to be defined for every combination of tracked/untracked arguments;
fortunately Julia's metaprogramming abilities made it simple for us to scale without the code becoming unwieldy.
We made a macro that first loops through all the \texttt{TrackedFloat} types that we wished to implement, and then looped through all of the operators that we wished to support.
This approached saved us over 10x the lines a non-macro approach would have required.\footnote{We implemented overrides for 645 function/type combinations. At roughly 5 lines per variant, this is 3225 lines of code. Our implementation at time of writing weighs in at 218 lines of code, which is \emph{including} whitespace and comments.}
This approach was inspired by \texttt{Sherlogs.jl}.\cite{kMilanklSherlogsJl2021}

\subsection{CSTG}
\label{s:cstg}

\FT{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced Stack Trace Graphs, or CSTGs~\cite{humphreySystematicDebuggingMethods2014} provide a way to visualize large amounts of stack traces in a compact form.
CSTG is a stand-alone tool and was not designed with \FT{} specifically in mind.
However, it pairs well with \FT{} and makes analyzing the log files easier.

CSTG takes a set of stack traces and makes a graph: for each stack trace, it creates a node for every frame, and then creates or increments an edge between two adjacent frames, as seen in \cref{fig:cstg_demo}.

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/cstg_static_diagram.png}
  \caption{CSTG makes a graph by connecting adjacent stack frames}
  \label{fig:cstg_demo}
\end{figure}

These graphs are meant to be read bottom-up: the bottom nodes are the roots where exceptional values manifested.
Higher nodes in the graph show function calls progressively closer to the top-level call that spawned them.
Thick lines and high numbers indicate flows with the highest volume leading to errors---though thin areas are still worth investigating.
CSTG provides a map to aid in contextualizing a stack trace in a log file.

@@

% HELP: @ben, I'm not sure what paper you mean by "Julia CPU"
Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In some regards they are \emph{more} susceptible because GPUs do not have the same exception-handling mechanisms that CPUs do.
Thus, tools to catch \fp{} exceptions on the GPU are essential.
\FT{} is constrained to CPU workloads; however, there is a companion tool called \GPUFPX{}~\cite{llsflg-hpdc-2023} that can help diagnose \fp{} exceptions on the GPU.
\GPUFPX{} can detect floating-point exceptions on the GPU that GPU code cannot do on its own.
\GPUFPX{} reports these exceptions back to the CPU for further analysis.

\FT{} and \GPUFPX{} can cover the entire computational stack, as seen in \cref{fig:ft_and_gpu_fpx}.
Work remains to tighten the integration between \FT{} and \GPUFPX{} to reduce manual work to employ both tools for heterogeneous computation.

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./fig/float_tracker_context_boxes.png}
  \caption{FloatTracker and GPU-FPX compliment each other}
  \label{fig:ft_and_gpu_fpx}
\end{figure}

@@

In this section we'll discuss some case studies where tools from \FlowFPX{} aided in finding or fixing \fp{} issues.
\Cref{s:sw} covers an ocean simulation that produces \NaN{}s in the output---a common and frustrating occurrence for many a scientist. \Cref{s:ode} puts the popular \texttt{OrdinaryDiffEq} solver library under a fuzzer to harden it against bad behavior in the presence of exceptional \fp{} values. \Cref{s:finch} fuzzes another library---this time a library for specifying partial differential equations (PDEs). Finally, in \cref{s:ocean}, we examine another ocean simulation library---this time the library runs on the GPU. This puts it out of reach of \FT{}, but \GPUFPX{} can handle it. In addition to the case studies that we did ourselves, \cref{s:safari} is an ``in the wild'' usage of \FT{} that helped researchers uncover the source of a \NaN{} in their own code without help from us.

@@

\texttt{ShallowWaters}~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019,klowerLowprecisionClimateComputing2021} simulates the flow of water over a seabed.
This is not a general shallow waters simulation library---it is designed primarily to showcase numerical methods involving 16-bit floats.

The Courant-Friedrichs-Lewy (``CFL'') parameter roughly describes the size of the time step to take in running the simulation.
Normally this value sits between 0 and 1---smaller values mean the simulation runs slower but with higher accuracy, whereas higher values mean the simulation takes bigger time steps and consequently runs faster, but results might be inaccurate.
The primary reason for problems arising comes from information not having enough time to propagate through the simulation before the next time step is taken.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{fig/sw_nan_cstg_clean.png}
  \caption{CSTG of \NaN{} generation logs from ShallowWaters.}
  \label{fig:sw_nan_cstg}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{fig/sw_inf_cstg_clean.png}
  \caption{CSTG of \Inf{} generation logs from the same run as~\cref{fig:sw_nan_cstg}.}
  \label{fig:sw_inf_cstg}
\end{figure}

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{pifont}

@@

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}
\newcommand{\GPUFPX}{GPU-FPX}
\newcommand{\FloatTracker}{FloatTracker}
\newcommand{\FT}{\FloatTracker}
\newcommand{\Fp}{Floating-point} % hyphen or no?
\newcommand{\fp}{floating-point} % hyphen or no?
\newcommand{\CSTG}{stack graph}
\newcommand{\CPP}{\code{C++}}
\newcommand{\Dendro}{\textsc{Dendro}}
\newcommand{\urlaccess}[2]{\url{#1}}
\newcommand{\Nan}{\code{NaN}}
\newcommand{\NaN}{\Nan}
\newcommand{\Inf}{\code{Inf}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zerocode}[1]{\zerowidth{\code{#1}}}
\newcommand{\genpropkill}{\emph{gen}-\emph{prop}-\emph{kill}}
\newcommand{\bigcheckmark}{\ding{51}}
\newcommand{\bigxmark}{\ding{53}}
\newcommand{\tblnext}{\(\Rightarrow\)}
\newcommand{\tblY}{\mbox{\bigcheckmark}}
\newcommand{\tblN}{\scalebox{0.8}{\bigxmark}}

@@

\begin{figure}[t]\centering
  \includegraphics[trim=0 0 0 60,clip,width=0.95\columnwidth]{fig/real_vs_fp.png}
  \caption{Floats are spread across the real number line}
  \label{f:real-vs-fp}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/genpropkill-outline.png}
  \caption{Gen, Prop, Kill: Lifetime of an exceptional value}
  \label{f:gpk}
\end{figure}

Unhandled exceptional values have a lifetime: they are born, or \emph{generated}, by some operation; they \emph{propagate} through other operations; and they either appear in the program output, go out of scope, or get \emph{killed} by a numeric operation.
\Cref{f:gpk} summarizes this \genpropkill{} process.
The gens and props are straightforward; see above for examples~(\cref{s:exnvalue}).
The kills often arise from numeric comparisons (\code{<}, \code{=}, etc.),
but exponents (\code{1{\string^}NaN}) and over-eager matrix
optimizations~\cite{ddghlllprr-correctness-2022} can kill exceptions as well.

@@

\FlowFPX{} is a toolkit for tracking down \fp{} exceptions.
The primary tools in this paper are \FT{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.

@@

\FT{} tracks exceptional values across their \genpropkill{} lifetime
and can fuzz code for vulnerabilities by injecting a \Nan{} or \Inf{}
as the result of an operation.
\FT{} is implemented in Julia and is available on JuliaHub:

\noindent\begin{center}\noindent\!\!\url{https://juliahub.com/ui/Packages/FloatTracker/dBXig}\!\!
\end{center}

@@

\begin{figure}[t]\centering
  \begin{tabular}{ccccc}
    Exn. Input? & \tblnext & Exn. Output? & = & Event       \\ \hline
    \tblN       & \tblnext &  \tblY{}     & = & {gen}  \\
    \tblY       & \tblnext &  \tblY{}     & = & {prop} \\
    \tblY       & \tblnext &  \tblN{}     & = & {kill} \\
  \end{tabular}
  \caption{How to classify operations that see exceptions}
  \label{fig:lifetime-class}
\end{figure}

\FT{} monitors \Nan{} and \Inf{} exceptions by overloading arithmetic
operations and logging key events.
When the input to an operation is exception-free but the output is
exceptional, the operation is a {gen} event~(\cref{fig:lifetime-class}).
When the input and output contain exceptions, the operation is a {prop}
event.
And when the input contains an exception but the output does not,
the operation is a {kill} event.
Put together, the log of all {gen}s, {prop}s, and {kill}s
sheds light of how various exceptions traveled across the program.
The logs also record the call context (stack trace) and the arguments to the operation as a starting point for debugging efforts.

\FT{} writes logs to three files: one for {gen} events, one for {prop}s, and one for {kill}s.
This way, discovering where a \Nan{} came from is a matter of sifting through the {gen} file.
Users can also lower the overhead of logging by turning it off for prop events.

The instrumentation works through custom \fp{} types: \code{TrackedFloat64},
\code{TrackedFloat32}, and \code{TrackedFloat16}.
Developers must opt in to \FT{} by wrapping numbers in a custom type.
From then on, tracking is automatic and extends transitively to all outputs
of tracked operations.
Multiplying a \code{Float64} value with a \code{TrackedFloat64}, for example, yields
a \code{TrackedFloat64} to continue the logging trail.

@@

Operator overloading gives \FT{} a powerful way to fuzz code from the inside out.
Each overloaded function serves as a hook where \FT{} can decide whether to observe
the operation or step in, discard the original result, and return an exception instead.
Injecting faults in a random way~\cite{hamlet1994random}, also known as fuzzing,
is a useful way to discover vulnerabilities in a large codebase.
Demmel et~al. propose essentially the same idea for BLAS and
LAPACK~\cite{ddghlllprr-correctness-2022}.

\FT{} exposes several parameters to let developers control the fuzzer:

\begin{itemize}
\item \texttt{odds::Int64} inject if \texttt{rand(odds) == 1}.
\item \texttt{n\_inject::Int64} upper bound on the number of \Nan{}s to inject.
\item \texttt{active::Bool} fuzz only when set to true.
\item \texttt{functions::Array{String}} limit fuzzing to the dynamic extent of the listed functions.
\item \texttt{libraries::Array{String}} limit fuzzing to functions from the following libraries
\end{itemize}

When fuzzing reveals an error, the next step is to craft a regression test to
guide repairs.
\FT{} therefore records the sequence of injections that it makes during a fuzzing
run and enables a replay of any recording after the fact.
Replay runs proceed deterministically so that developers can harden
their code and check that the fixes remove the error.

@@

\FT{} takes advantage of Julia's operator overloading to track exceptions and
fuzz for vulnerabilities.
For example, below is a variant of \code{+} that is overloaded
for \code{TrackedFloat64}.
It calls the basic \code{+} (or injects a \Nan{}) and checks for exceptions
before returning:
This lets \FT{} intercept all \fp{} operations involving \texttt{TrackedFloat} types.

@@

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \FT{} uses a metaprogramming technique
adapted from \texttt{Sherlogs.jl}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

\subsection{Stack Graphs}
\label{s:cstg}

\FT{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced stack trace graphs (CSTGs or \CSTG{}s for short) provide a way to visualize large amounts of stack traces in a compact form~\cite{humphreySystematicDebuggingMethods2014}.
This techinque pairs well with \FT{} and makes analyzing the log files easier.

\Cref{fig:cstg_demo} illustrates the construction of a \CSTG{} from a collection of stack traces.
Each trace on the left contributes nodes and edges to the graph on the right.
Repeated edges get emphasized with darker lines and larger counts.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{./fig/cstg_static_diagram.png}
  \caption{From stack traces (left) to \CSTG{} (right)}
  \label{fig:cstg_demo}
\end{figure}

Reading bottom-up, a \CSTG{} based on the \emph{gen} events in a program
highlights the contexts that frequently produced exceptional values.
Using~\cref{fig:cstg_demo} as an example, \code{Frame D} produced
three exceptions, two of which arose under \code{Frame C}.
In a large program with many exceptions, \CSTG{}s offer a way to prioritize
debugging efforts: go for the heavily-trodden paths first.

@@

Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In fact, GPU programs are worse off because they lack exception-handling mechanisms from the CPU world~\cite{llg-soap-2022}.
Since \FT{} instruments Julia programs, it cannot help directly;
however, the companion tool \GPUFPX{} instruments GPU kernels to detect and report
\fp{} exceptions~\cite{llsflg-hpdc-2023}.
Together, \FT{} and \GPUFPX{} provide insights for accelerated programs.

@@

\FlowFPX{} has helped to debug exceptions and fuzz for issues in a variety
of settings, some synthetic and some realistic.
The case studies include a shallow water simulation, the
\texttt{OrdinaryDiffEq} solver, and a Bayesian inference library.

@@

\texttt{ShallowWaters.jl} simulates the flow of water over a
seabed~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019}.
The library has dozens of parameters that a scientist can experiment with.
One notable parameter is the Courant-Friedrichs-Lewy (CFL) number, which roughly describes the size of the time step to take in running the simulation.
A small CFL number makes the simulation run slowly, but accurately; a large number speeds it up but loses precision
because the system does not get enough time to propagate information.

Normal CFL values range between zero and one.
X

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.96\columnwidth]{fig/sw_nan_cstg_clean.png}
  \caption{CSTG of \NaN{} generation logs from ShallowWaters.}
  \label{fig:sw_nan_cstg}
\end{figure}

% \begin{figure}[t]
%   \centering
%   \includegraphics[width=3in]{fig/sw_inf_cstg_clean.png}
%   \caption{CSTG of \Inf{} generation logs from the same run as~\cref{fig:sw_nan_cstg}.}
%   \label{fig:sw_inf_cstg}
% \end{figure}
